The 2- and 3-point surface correlation functions calculations: from novel exact continuous approach to improving methodology for discrete images

### 2023-08-04 paper.tex
Replase Fss plots for disks
Update plots (ellipses)
Update plots (blobs)
Last edits from Kirill

\begin{abstract}
  Surface correlation functions are extremely useful descriptors of interfaces
  and in this role can be used in material science, rock and soil physics, food
  engineering and other numerous areas. However, their widespread usage was
  hampered by insufficiently robust computation methodologies. In this
  contribution we elaborate on recent works in surface function evaluation and
  present a novel exact algorithm to compute surface-surface ($F_{ss}$) and
  surface-surface-surface ($F_{sss}$) functions for smooth boundary sets without
  discretization. This allows us to test digital computational approach on
  geometries beyond those available as analytical solutions. Moreover, we
  propose a better edge detection filter that allowed to improve the quality of
  assessment of surface correlation functions, both surface-surface and
  surface-void ($F_{sv}$), for digital images. The code for continuous and
  discrete computational frameworks is freely available and is now the part of
  the more general CorrelationFunctions.jl package.
\end{abstract}

@@

In majority, if not all, cases mentioned above, surface functions - namely,
$F_{ss}$ and $F_{sv}$ - were underrepresented due to our inability to compute
them in a robust way. While known for a long time \cite{dietrich1995scattering}, only
somewhat crude approximate approaches \cite{seaton1986spatial} were developed
until Ma and Torquato laid foundation for such a robust computational
methodology \cite{ma2018SS}. They proposed an elegant solution for infinite
resolution Gaussian random fields, or "continuous" approach. They also assumed
that the same approach can be readily applied to discrete experimental images if
grey-scale values are thresholded similarly to continuous Gaussian
field. Samarin et al. \cite{Samarin} argued that experimental images needs
to be segmented first using current cutting edge techniques \cite{NNseg} and
proposed "discrete" approach. The latter allowed obtaining surface functions
closer to analytical solutions for overlapping disks/spheres and to suppress
perturbation, in particular for $F_{ss}$. The perturbation in Ma and Torquato
method resulted from $1/cos(\theta)$ computations during random line segment
placement -- this leaves a room for improvements. While Samarin et al. obtained
better convergence with analytical solution, this was demonstrated only for
overlapping disks image of limited size (that can potentially introduce some
differences between computations and analytical solution for infinite size
structure). Moreover, good agreement was achieved by manually fitting the values
of edge-detecting filter that could lack accuracy for arbitrary structure
different from overlapping disks. In other words, we still lack robustness in
surface CFs evaluation due to inability to directly compare "discrete" versus
"continuous" methodology, and verify both approaches for arbitrary geometries.

@@

Analytical approaches have been proposed to compute only n-point functions $S_n$
\cite{PhysRevB.42.4453,10.1063/1.4865966}. but studying n-point surface
correlations faces lack of analytical solutions. Availability of analytical
solutions for correlation functions holds significant importance and serves as a
desirable benchmark for evaluating and validating numerical algorithms. However,
a comprehensive computational framework for calculating higher-order correlation
functions in heterogeneous media is still lacking. In their study
\cite{malmir2018}, the authors proposed an algorithm for computing third-order
correlation functions and compared its performance with existing analytical
solutions and asymptotic cases. Although the algorithm demonstrates good
performance for the computation of S3 functions, the authors themselves
acknowledge that there is room for further improvement, particularly in the case
of surface correlation functions. To address this limitation, they suggest
employing a method similar to the one used in the work of Ma and Torquato
\cite{ma2018SS}, where a digitized image is converted into a scalar field. In
our current study, we advance beyond these findings and significantly enhance
the numerical solution for the three-point correlation function of surfaces by
incorporating and developing ideas from the work of Samarin et al
\cite{Samarin}.

In this contribution we continue to build upon foundational work of Ma and
Torquato \cite{ma2018SS} and recent findings of Samarin et al. \cite{Samarin} to
develop a robust and efficient approach to compute surface correlation functions
from digital 2D and 3D images. In particular, here our aim is to establish a
direct verification of "discrete" methodology (that is easily parallelizable on
CPU and GPU architectures) using exact "continuous" approach for arbitrary
structures (still described by some continuous function).

@@

Now we define two disjoint sets in a homogeneous porous medium: a set of solid
phase $S$ and a set of void phase $V$. Surface-surface correlation function
$F_{ss}$ and surface-void correlation function $F_{sv}$ are then defined as
follows: \cite{Torquato_book}
\begin{align}
  F_{ss}(\bm{r}) &= \langle M_S(\bm{x}) M_S(\bm{x} + \bm{r}) \rangle \label{eq:fss} \\
  F_{sv}(\bm{r}) &= \langle M_S(\bm{x}) \chi_V(\bm{x} + \bm{r}) \rangle
  \label{eq:fsv}
\end{align}
Here the brackets $\langle \dots \rangle$ denote ensemble average. If the medium
under consideration is also isotropic, the argument in functions $F_{ss}$ and
$F_{sv}$ can be replaced with a scalar $r = |\bm{r}|$:
\begin{align}
  F_{ss}(r) &= F_{ss}(\bm{r}) \\
  F_{sv}(r) &= F_{sv}(\bm{r})
\end{align}
In case correlation functions are computed in predefined directions, scalar
values can be also used for anisotropic structures
\cite{10.1063/1.4867611,EPL1}. These functions are defined for media with
dimensionality of 2 and higher. $F_{sv}$ is a physical quantity inversely
proportional to length and $F_{ss}$ is inversely proportional to surface
area. SI units of measurement for these functions are $m^{-1}$ for $F_{sv}$
and $m^{-2}$ for $F_{ss}$. In practice non-standard units of measurement can be
used instead.

Another function studied in this paper is three-point surface-surface-surface
correlation function $F_{sss}$. It is defined as follows:
\begin{equation}
  F_{sss}(\bm{r_1}, \bm{r_2}) = \langle M_S(\bm{x}) M_S(\bm{x} + \bm{r_1})
  M_S(\bm{x} + \bm{r_2}) \rangle
  \label{eq:fsss}
\end{equation}
This function is defined for media with dimensionality 3 or higher and is a
physical quantity inversely proportional to volume. SI units for $F_{sss}$ are
$m^{-3}$, but relative units can be used in practice. For three-dimensional media
$F_{sss}$ is a function of six scalar arguments:
$\bm{r_1} = (r_{1x}, r_{1y}, r_{1z})$ and $\bm{r_2} = (r_{2x}, r_{2y}, r_{2z})$.
However, if the medium under consideration is isotropic, only three scalar
arguments $r_1 = |\bm{r_1}|$, $r_2 = |\bm{r_2}|$ and $r_3 = |\bm{r_1} - \bm{r_2}|$
are required. The last argument can be replaced with an angle $\theta$ between
vectors $\bm{r_1}$ and $\bm{r_2}$:
\begin{equation}
  F_{sss}(\bm{r_1}, \bm{r_2}) = F_{sss}(r_1, r_2, r_3) = F_{sss}(r_1, r_2, \theta)
\end{equation}

@@

However, the $H'$ filter has some drawbacks as compared to $H$. Firstly, $H$ has
faster convergence with increase of image resolution. Secondly, $H'$ behaves
worse than $H$ in points of discontinuity of $F_{ss}$ as can be seen on
\cref{fig:fss-disk}. To explore such effects we appeal to the parameter
$C_\alpha$ \cite{Samarin} developed to determine if an image is suitable for
robust assessment of surface correlation functions. To introduce this parameter,
consider a continuos band-limited signal $f(t)$ with support in frequency domian
$[-\omega_0, \omega_0]$. The $L^2$ norm of this signal is
\begin{equation}
  \|f\| = \sqrt{\int_{-\omega_0}^{\omega_0} |\hat{f}(\omega)|^2 d\omega}
\end{equation}
where $\hat{f}(\omega)$ is a Fourier transform of $f(t)$. The idea behind the
parameter is to determine how much low-frequency domain contributes to the total
power of a signal. $C_\alpha$ parameter is defined as
\begin{equation}
  \begin{aligned}
    C_\alpha &= \sqrt{\frac{\int_{-\alpha\omega_0}^{\alpha\omega_0}
        |\hat{f}(\omega)|^2 d\omega}{\int_{-\omega_0}^{\omega_0}
        |\hat{f}(\omega)|^2 d\omega}} \\
    \alpha &\in [0, 1]
  \end{aligned}
  \label{eq:parameter}
\end{equation}
For two- and three-dimensional discrete images $f(t)$ is replaced with the input
array $A$ minus its mean, integrals in \cref{eq:parameter} are replaced with
double and triple sums respectively and $\omega_0$ is equal to the Nyquist
frequency for the image. Our empirical criterion says that the image is suitable
for evaluation of surface correlation functions if
\begin{equation}
  C_{0.5} > 0.97
\end{equation}
This criterion works for both filters $H$ and $H'$, but $H$ results in smaller
error when $C_{0.5}$ is slightly less than 0.97. For example, consider images of
a disk used for calculation of $F_{ss}$ on \cref{fig:fss-disk}. On
\cref{fig:error} we provide values of $C_{0.5}$ for this image (for variety of
resolutions) and relative errors in $F_{ss}$ (\cref{eq:error}) for both
filters. It can be seen that $H$ gives twice the lower error than $H'$ when used
for an image with $C_{0.5} \approx 0.82$. In short, for images of acceptable
quality the $H'$ filter is superior.

@@

>>>>

\begin{abstract}
  Surface correlation functions are extremely useful descriptors of interfaces
  and in this role can be used in material science, rock and soil physics, food
  engineering and other numerous areas. However, their widespread usage was
  hampered by insufficiently robust computation methodologies. In this
  contribution we elaborate on recent works in surface function evaluation and
  present a novel exact algorithm to compute surface-surface ($F_{ss}$) and
  surface-surface-surface ($F_{sss}$) functions for smooth boundary sets without
  discretization. This allows us to test digital computational approach on
  geometries beyond those available as analytical solutions. Moreover, we
  propose a better edge detection filter that allowed to improve the quality of
  assessment of surface correlation functions, both surface-surface ($F_{ss}$) and
  surface-void ($F_{sv}$), for digital images. The code for continuous and
  discrete computational frameworks is freely available and is now the part of
  the more general CorrelationFunctions.jl package.
\end{abstract}

@@

In majority, if not all, cases mentioned above, surface functions - namely,
$F_{ss}$ and $F_{sv}$ - were underrepresented due to our inability to compute
them in a robust way. While known for a long time \cite{dietrich1995scattering}, only
somewhat crude approximate approaches \cite{seaton1986spatial} were developed
until Ma and Torquato laid foundation for a robust computational
methodology \cite{ma2018SS}. They proposed an elegant solution for infinite
resolution Gaussian random fields, or "continuous" approach. They also assumed
that the same approach can be readily applied to discrete experimental images if
grey-scale values are thresholded similarly to continuous Gaussian
field. Samarin et al. \cite{Samarin} argued that experimental images needs
to be segmented first using current cutting edge techniques \cite{NNseg} and
proposed "discrete" approach. The latter allowed obtaining surface functions
closer to analytical solutions for overlapping disks/spheres and to suppress
noise, in particular for $F_{ss}$. The noise in Ma and Torquato
method resulted from $1/cos(\theta)$ computations during random line segment
placement -- this leaves a room for improvements. While Samarin et al. obtained
better convergence with analytical solution, this was demonstrated only for
overlapping disks image of limited size (that can potentially introduce some
differences between computations and analytical solution for infinite size
structure). Moreover, good agreement was achieved by manually fitting the values
of edge-detecting filter that could lack accuracy for arbitrary structure
different from overlapping disks. In other words, we still lack robustness in
surface CFs evaluation due to inability to directly compare "discrete" versus
"continuous" methodology, and verify both approaches for arbitrary geometries.

@@

Analytical approaches have been proposed to compute only n-point functions $S_n$
\cite{PhysRevB.42.4453,10.1063/1.4865966}. but studying n-point surface
correlations faces lack of analytical solutions. Availability of analytical
solutions for correlation functions holds significant importance and serves as a
desirable benchmark for evaluating and validating numerical algorithms. However,
a comprehensive computational framework for calculating higher-order correlation
functions in heterogeneous media is still lacking. In their study
\cite{malmir2018}, the authors proposed an algorithm for computing third-order
correlation functions and compared its performance with existing analytical
solutions and asymptotic cases. Although the algorithm demonstrates good
performance for the computation of $S_3$ functions, the authors themselves
acknowledge that there is room for further improvement, particularly in the case
of surface correlation functions. To address this limitation, they suggest
employing a method similar to the one used in the work of Ma and Torquato
\cite{ma2018SS}, where a digitized image is converted into a scalar field. In
our current study, we advance beyond these findings and significantly enhance
the numerical solution for the three-point correlation function of surfaces.

All in all, we continue to build upon foundational work of Ma and
Torquato \cite{ma2018SS} and recent findings of Samarin et al. \cite{Samarin} to
develop a robust and efficient approach to compute surface correlation functions
from digital 2D and 3D images. In particular, here our aim is to establish a
direct verification of "discrete" methodology (that is easily parallelizable on
CPU and GPU architectures) using exact "continuous" approach for arbitrary
structures (still described by some continuous function).

@@

Now we define two disjoint sets in a homogeneous porous medium: a set of solid
phase $S$ and a set of void phase $V$. Surface-surface correlation function
$F_{ss}$ and surface-void correlation function $F_{sv}$ are then defined as
follows: \cite{Torquato_book}
\begin{align}
  F_{ss}(\bm{r}) &= \langle M_S(\bm{x}) M_S(\bm{x} + \bm{r}) \rangle \label{eq:fss} \\
  F_{sv}(\bm{r}) &= \langle M_S(\bm{x}) \chi_V(\bm{x} + \bm{r}) \rangle
  \label{eq:fsv}
\end{align}
Here the brackets $\langle \dots \rangle$ denote ensemble average. If the medium
under consideration is also isotropic, the argument in functions $F_{ss}$ and
$F_{sv}$ can be replaced with a scalar $r = |\bm{r}|$:
\begin{align}
  F_{ss}(r) &= F_{ss}(\bm{r}) \\
  F_{sv}(r) &= F_{sv}(\bm{r})
\end{align}
In case correlation functions are computed in predefined directions, scalar
values can be also used for anisotropic structures
\cite{10.1063/1.4867611,EPL1}. $F_{sv}$ is a physical quantity inversely
proportional to length and $F_{ss}$ is inversely proportional to surface
area. SI units of measurement for these functions are $m^{-1}$ for $F_{sv}$
and $m^{-2}$ for $F_{ss}$. In practice non-standard units of measurement can be
used instead.

Another function studied in this paper is three-point surface-surface-surface
correlation function $F_{sss}$. It is defined as follows:
\begin{equation}
  F_{sss}(\bm{r_1}, \bm{r_2}) = \langle M_S(\bm{x}) M_S(\bm{x} + \bm{r_1})
  M_S(\bm{x} + \bm{r_2}) \rangle
  \label{eq:fsss}
\end{equation}
SI units for $F_{sss}$ are $m^{-3}$, but relative units can be used in practice. For three-dimensional media
$F_{sss}$ is a function of six scalar arguments:
$\bm{r_1} = (r_{1x}, r_{1y}, r_{1z})$ and $\bm{r_2} = (r_{2x}, r_{2y}, r_{2z})$.
However, if the medium under consideration is isotropic, only three scalar
arguments $r_1 = |\bm{r_1}|$, $r_2 = |\bm{r_2}|$ and $r_3 = |\bm{r_1} - \bm{r_2}|$
are required. The last argument can be replaced with an angle $\theta$ between
vectors $\bm{r_1}$ and $\bm{r_2}$:
\begin{equation}
  F_{sss}(\bm{r_1}, \bm{r_2}) = F_{sss}(r_1, r_2, r_3) = F_{sss}(r_1, r_2, \theta)
\end{equation}

@@

However, the $H'$ filter has some drawbacks as compared to $H$. Firstly, $H$ has
faster convergence with increase of image resolution. Secondly, $H'$ behaves
worse than $H$ in points of discontinuity of $F_{ss}$ as can be seen on
\cref{fig:fss-disk}. To explore such effects we appeal to the parameter
$C_\alpha$ \cite{Samarin} developed to determine if an image is suitable for
robust assessment of surface correlation functions. To introduce this parameter,
consider a continuos band-limited signal $f(t)$ with support in frequency domian
$[-\omega_0, \omega_0]$. The $L^2$ norm of this signal is:
\begin{equation}
  \|f\| = \sqrt{\int_{-\omega_0}^{\omega_0} |\hat{f}(\omega)|^2 d\omega}
\end{equation}
where $\hat{f}(\omega)$ is a Fourier transform of $f(t)$. The idea behind the
parameter is to determine how much low-frequency domain contributes to the total
power of a signal. $C_\alpha$ parameter is defined as:
\begin{equation}
  \begin{aligned}
    C_\alpha &= \sqrt{\frac{\int_{-\alpha\omega_0}^{\alpha\omega_0}
        |\hat{f}(\omega)|^2 d\omega}{\int_{-\omega_0}^{\omega_0}
        |\hat{f}(\omega)|^2 d\omega}} \\
    \alpha &\in [0, 1]
  \end{aligned}
  \label{eq:parameter}
\end{equation}
For two- and three-dimensional discrete images $f(t)$ is replaced with the input
array $A$ minus its mean, integrals in \cref{eq:parameter} are replaced with
double and triple sums respectively and $\omega_0$ is equal to the Nyquist
frequency for the image. Our empirical criterion says that the image is suitable
for evaluation of surface correlation functions if:
\begin{equation}
  C_{0.5} > 0.97
\end{equation}
This criterion works for both filters $H$ and $H'$, but $H$ results in smaller
error when $C_{0.5}$ is slightly less than 0.97. For example, consider images of
a disk used for calculation of $F_{ss}$ on \cref{fig:fss-disk}. On
\cref{fig:error} we provide values of $C_{0.5}$ for this image (for variety of
resolutions) and relative errors in $F_{ss}$ (\cref{eq:error}) for both
filters. It can be seen that $H$ gives twice the lower error than $H'$ when used
for an image with $C_{0.5} \approx 0.82$. In short, for images of acceptable
quality the $H'$ filter is superior.

@@

\section{Acknowledgements}
This research was supported by the Russian Science Foundation grant 23-74-00061. Collaborative effort of the authors is within the FaT iMP 
(Flow and Transport in Media with Pores) research group (www.porenetwork.com). We thank our colleague Konstantin Romanenko for suggestions 
and administrative work.
